---
output:
  pdf_document: default
  html_document: default
---
# Selected MAGs

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(distillR)
library(ape)
library(vegan)
library(ggrepel)
library(ggh4x)
library(broom)

```

```{r load_data, message=FALSE, warning=FALSE}
gene_annotations <- read_tsv("data/gene_annotations.tsv.xz") %>% 
  filter(gene!="gene") %>% 
  mutate(ec=str_c("[EC:",ec,"]")) %>% 
  mutate(genome=substr(gene, 1, 7))

genome_metadata <- read_csv("data/genome_metadata.csv")
```

```{r distill_annotations, message=FALSE, warning=FALSE}
gene_distillation <- distill(gene_annotations,GIFT_db, genomecol = 14, annotcol = c(5,6), verbosity=F)
gift_elements <- to.elements(gene_distillation,GIFT_db)
```

```{r distill_plot, message=FALSE, warning=FALSE, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
gift_elements %>%
    as.data.frame() %>%
    rownames_to_column(var="genome") %>%
    pivot_longer(!genome,names_to="trait",values_to="gift") %>%
    inner_join(genome_metadata,by="genome") %>% 
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=genome,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_nested(functionid ~ group + farm, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 0)) + 
        labs(y="Traits",x="Samples",fill="GIFT")
```


## Functional differences

Ideally, a mixed effects modelling should be used, but susceptible MAGs are too few for any reasonable modelling.

```{r function_test, message=FALSE}
functional_differences <- gift_elements %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  inner_join(genome_metadata,by="genome") %>% 
  pivot_longer(-c(genome, group, farm), names_to = "trait", values_to = "value") %>%
  nest(data = -trait) %>%                                              
  mutate(
    fit  = map(data, ~lm(value ~ group, data = .x)),
    tid  = map(fit, tidy) 
  ) %>% 
  unnest(tid) %>% 
  filter(term == "groupsusceptible") %>%
  mutate(p_value_adj = p.adjust(p.value, method = "bonferroni")) %>% 
  select(id   = trait,
         estimate,
         p.value,
         p_value_adj) %>% 
  arrange(p.value)

functional_differences %>% 
  filter(p.value<0.05)
```

None of the GIFTs yield significant differences after Bonferroni adjustment.

```{r function_test_plot, message=FALSE}
gift_elements %>% 
  as.data.frame() %>% 
  select(functional_differences %>% filter(p.value<0.05) %>% pull(id)) %>% 
  rownames_to_column(var="genome") %>% 
  pivot_longer(!genome,names_to = "trait", values_to = "value") %>% 
  inner_join(genome_metadata,by="genome") %>% 
  ggplot(aes(x=group, y=value, group=group))+
    geom_boxplot() +
    geom_jitter() +
    facet_grid(. ~ trait, scales="free",space="free")

```

## Functional ordination 

```{r function_ordination_pcoa, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
gift_pcoa <- gift_elements %>%
    as.data.frame() %>%
    vegdist(method="euclidean") %>%
    pcoa()

gift_pcoa_rel_eigen <- gift_pcoa$values$Relative_eig[1:10]


# Get genome positions
gift_pcoa_vectors <- gift_pcoa$vectors %>% #extract vectors
  as.data.frame() %>% 
  select(Axis.1,Axis.2) # keep the first 2 axes

gift_pcoa_eigenvalues <- gift_pcoa$values$Eigenvalues[c(1,2)]

gift_pcoa_gifts <- cov(gift_elements, scale(gift_pcoa_vectors)) %*% diag((gift_pcoa_eigenvalues/(nrow(gift_elements)-1))^(-0.5)) %>%
  as.data.frame() %>% 
  rename(Axis.1=1,Axis.2=2) %>% 
  rownames_to_column(var="label") %>% 
  #get function summary vectors
  mutate(func=substr(label,1,3)) %>% 
  group_by(func) %>% 
  summarise(Axis.1=mean(Axis.1),
            Axis.2=mean(Axis.2)) %>% 
  rename(label=func) %>% 
  filter(!label %in% c("S01","S02","S03"))

```

```{r function_ordination_pcoa_plot, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

scale <- 15 # scale for vector loadings
gift_pcoa_vectors %>% 
  rownames_to_column(var="genome") %>% 
  inner_join(genome_metadata,by="genome") %>% 
  group_by(farm) %>%
  mutate(x_cen = mean(Axis.1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(Axis.2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot() +
      #genome positions
      #scale_color_manual(values=order_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color=group, shape=farm), alpha=0.9, size=4) +
      geom_segment(aes(x = x_cen, y = y_cen, xend = Axis.1, yend = Axis.2, group=farm), alpha = 0.9, color="grey", show.legend = FALSE) +
      #scale_color_manual(values=phylum_colors) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen[1]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA1")
            ) +
     scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen[2]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA2")
            ) +
    geom_label_repel(data = gift_pcoa_gifts,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
    theme_minimal() + 
    theme(legend.position = "none")

```

There is a huge effect of the farm in the functional profile of the MAGs. The lines connect all the MAGs from each farm (dot and triangle).